---
title: "Chapter 4 Solutions"
author: "Benjamin Wee"
output: 
  html_document:
    theme: cosmo
    code_folding: show
---

```{r message=FALSE, warning=FALSE}
library(rethinking)
```

# Chapter 4
## Easy

$$
y_i\sim Normal(\mu,\sigma) \\
\mu\sim Normal(0, 10) \\
\sigma\sim Uniform(0, 10)
$$

### 4E1
$y_i\sim Normal(\mu,\sigma)$ is the likelihood

### 4E2
2 parameters in the posterior, $\mu$ and $\sigma$

### 4E3
$$
Pr(\mu,\sigma|y_i) = \frac{Pr(y_i|\mu,\sigma) Pr(\mu) Pr(\sigma)}{Pr(y_i)} \\
= \frac{\Pi_i Normal(y_i|\mu,\sigma) Normal(\mu | 0, 10) Uniform(\sigma|0, 10)}{\int\int\Pi_i Pr(y_i)Normal(y_i|\mu,\sigma) Normal(\mu | 0, 10) Uniform(\sigma|0, 10)d\mu d\sigma}
$$

### 4E4
$\mu_i = \alpha + \beta x_i$ is the linear model

### 4E5
3 parameters in the posterior, $\alpha$, $\beta$ and $\sigma$


## Medium
$$
y_i\sim Normal(\mu,\sigma) \\
\mu\sim Normal(0, 10) \\
\sigma\sim Uniform(0, 10)
$$

### 4M1
```{r}
sample_mu <- rnorm(1e4, 0, 10)
sample_sigma <- runif(1e4, 0, 10)
y <- rnorm(1e4, sample_mu, sample_sigma)
```

### 4M2
```{r}
model <- alist(y ~ dnorm(mu, sigma),
               mu ~ dnorm(0, 10),
               sigma ~ dunif(0, 10))
```

### 4M3
$$
y_i\sim Normal(\mu,\sigma) \\
\mu = \alpha + \beta x_i \\
\alpha\sim Normal(0, 50) \\
\beta\sim Normal(0, 10) \\
\sigma\sim Uniform(0, 50)
$$

### 4M4
$$
Height_i\sim Normal(\mu,\sigma) \\
\mu = \alpha + \beta time_i \\
\alpha\sim Normal(160, 100) \\
\beta\sim Normal(0, 10) \\
\sigma\sim Uniform(0, 50)
$$
Wide prior for intercept. Height does not have a linear relationship with age/time. Most likely some concave function.

If adolescent students, probably some growth across teenage years. 

### 4M5
$$
Height_i\sim Normal(\mu,\sigma) \\
\mu = \alpha + \beta time_i \\
\alpha\sim Normal(120, 100) \\
\beta\sim Normal(2, 10) \\
\sigma\sim Uniform(0, 50)
$$

Mean of alpha becomes 120 to reflect average height of first years. Positive mean to slope to reflect overall growth for a year. 

### 4M6
$$
Height_i\sim Normal(\mu,\sigma) \\
\mu = \alpha + \beta time_i \\
\alpha\sim Normal(120, 100) \\
\beta\sim Normal(2, 10) \\
\sigma\sim Uniform(0, 70)
$$

Set upper bound for variance of height to be 70. Higher than observed upper bound to account for potential variation that was not previously observed (although should probably not have uniform weight on values above 64).

## Hard

### 4H1

Using code model from 4.38 

```{r message=FALSE, warning=FALSE}
library(rethinking)
data(Howell1)
d <- Howell1

# Define univariate linear model
m4.3 <- map(
  alist(
    height ~ dnorm( mu , sigma ) ,
    mu <- a + b*weight ,
    a ~ dnorm( 156 , 100 ) ,
    b ~ dnorm( 0 , 10 ) ,
    sigma ~ dunif( 0 , 50 )
) ,
data=d)

missing_data <- as.data.frame(c(46.95, 43.72, 64.78, 32.59, 54.63))
names(missing_data) <- "weight"

# Simulate expected (mean) heights 
mu_missing <- link(m4.3, data = missing_data)
mu_missing_mean <- apply(mu_missing, 2, mean)
mu_missing_hpdi <- apply(mu_missing, 2, HPDI, prob = 0.89)

print(rbind(mu_missing_mean, mu_missing_hpdi))
```

### 4H2
```{r}
# 4H2a
library(rethinking)
data(Howell1)
d <- Howell1[Howell1$age < 18, ]

model <- map(
  alist(
    height ~ dnorm( mu , sigma ) ,
    mu <- a + b*weight ,
    a ~ dnorm( 100 , 100 ) ,
    b ~ dnorm( 0 , 10 ) ,
    sigma ~ dunif( 0 , 50 )
) ,
data=d)

precis(model)
```

```{r}
# 4H2b
# Define sequence of weights
weight.seq <- seq(from = 1, to = 50)

# Calculate mu
mu <- link(model, data = data.frame(weight = weight.seq))

# Calculate mu.mean
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply(mu, 2, HPDI, prob = 0.89)

# Simulate heights
sim.height <- sim(model, data = list(weight=weight.seq))
height.HPDI <- apply(sim.height, 2, HPDI, prob = 0.89)

# Plot 
plot(height ~ weight, data = d, col = col.alpha(rangi2, 0.5))
lines(weight.seq, mu.mean)
shade(mu.HPDI, weight.seq)
shade(height.HPDI, weight.seq)
```

#### 4H3c
Plotted data does not really have a linear relationship. Maybe add a polynomial regressor to account for concavity in the relationship between height and weight. 

### 4H3
```{r}
library(rethinking)
data(Howell1)
d <- Howell1

log_model <- map(
  alist(
    height ~ dnorm( mu , sigma ) ,
    mu <- a + b*log(weight),
    a ~ dnorm( 178 , 100 ) ,
    b ~ dnorm( 0 , 100 ) ,
    sigma ~ dunif( 0 , 50 )
) ,
data=d)

precis(log_model)
```

```{r}
#### 4H3b
# Define sequence of weights
weight.seq <- seq(from = 1, to = 70)

# Extract samples from model
posterior <- extract.samples(log_model, n = 1e5)

# Calculate mu
mu_link <- function(weight) posterior$a + posterior$b * log(weight)
mu <- sapply(X = weight.seq, FUN = mu_link)

# Calculate mu.mean
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply(mu, 2, HPDI, prob = 0.97)

# Simulate heights
height.link <- function(weight) rnorm(n = nrow(posterior), mean = mu_link(weight), sd = posterior$sigma)
height.samples <- sapply(weight.seq, FUN = height.link)
height.HPDI <- apply(height.samples, 2, HPDI, prob = 0.97)

plot( height ~ weight , data=Howell1,
col=col.alpha(rangi2,0.4) )
lines(weight.seq, mu.mean)
shade(mu.HPDI, weight.seq)
shade(height.HPDI, weight.seq)
```



