---
title: "Chapter 8 - Stan"
author: "Benjamin Wee"
output: html_document
---

# Medium

## 8M1
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rethinking)
data('rugged', package = 'rethinking')

# Log outcom
rugged$log_gdp <- log(rugged$rgdppc_2000)

# Filter out countries without GDP data
rugged_gdp <- rugged %>% filter(!is.na(rgdppc_2000))
```

Uniform Prior
```{stan output.var=mm1_1}
data {
  int N;
  vector[N] rugged;
  vector[N] cont_africa;
  vector[N] log_gdp;
}
parameters{
  real a;
  real bR;
  real bA;
  real bAR;
  real<lower=0, upper=10> sigma;
}
model{
  vector[N] gamma;
  vector[N] mu;
  
  for (i in 1:N){
  gamma[i] = bR + bAR * cont_africa[i];
  mu[i] = a + gamma[i] * rugged[i] + bA * cont_africa[i];
  }
  
  target += normal_lpdf(log_gdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 10);
  target += normal_lpdf(bAR | 0, 10);
  target += normal_lpdf(bA | 0, 10);
}
generated quantities{
  vector[N] log_lik;
  {
  vector[N] mu;
  vector[N] gamma;
  for (n in 1:N){
  gamma[n] = bR + bAR*cont_africa[n];
  mu[n] = a + gamma[n] * rugged[n] + bA * cont_africa[n];
  log_lik[n] = normal_lpdf(log_gdp[n] | mu[n], sigma);
  }
  }
}
```

Exponential
```{stan output.var=mm1_2}
data {
  int N;
  vector[N] rugged;
  vector[N] cont_africa;
  vector[N] log_gdp;
}
parameters{
  real a;
  real bR;
  real bA;
  real bAR;
  real<lower=0> sigma;
}
model{
  vector[N] gamma;
  vector[N] mu;
  
  for (i in 1:N){
  gamma[i] = bR + bAR * cont_africa[i];
  mu[i] = a + gamma[i] * rugged[i] + bA * cont_africa[i];
  }
  
  target += normal_lpdf(log_gdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 10);
  target += normal_lpdf(bAR | 0, 10);
  target += normal_lpdf(bA | 0, 10);
  target += exponential_lpdf(sigma | 1);
}
generated quantities{
  vector[N] log_lik;
  {
  vector[N] mu;
  vector[N] gamma;
  for (n in 1:N){
  gamma[n] = bR + bAR*cont_africa[n];
  mu[n] = a + gamma[n] * rugged[n] + bA * cont_africa[n];
  log_lik[n] = normal_lpdf(log_gdp[n] | mu[n], sigma);
  }
  }
}
```

Sample
```{r}
data_m1 <- list(
  N = NROW(rugged_gdp),
  rugged = rugged_gdp$rugged,
  cont_africa= rugged_gdp$cont_africa,
  log_gdp = rugged_gdp$log_gdp
)

fitm_m1_1 <- sampling(mm1_1, data = data_m1, iter = 20000, chains = 4, cores = 4)
fitm_m1_2 <- sampling(mm1_2, data = data_m1, iter = 20000, chains = 4, cores = 4)

```

```{r}
fitm_m1_1
```

```{r}
fitm_m1_2
```

Mean and overall shape of sigma posterior about the same. Priors are quite weak.

## 8M2

Exponential
```{stan output.var=mm2_1}
data {
  int N;
  vector[N] rugged;
  vector[N] cont_africa;
  vector[N] log_gdp;
}
parameters{
  real a;
  real bR;
  real bA;
  real bAR;
  real<lower=0> sigma;
}
model{
  vector[N] gamma;
  vector[N] mu;
  
  for (i in 1:N){
  gamma[i] = bR + bAR * cont_africa[i];
  mu[i] = a + gamma[i] * rugged[i] + bA * cont_africa[i];
  }
  
  target += normal_lpdf(log_gdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 10);
  target += normal_lpdf(bAR | 0, 10);
  target += normal_lpdf(bA | 0, 10);
  target += exponential_lpdf(sigma | 0.1);
}
generated quantities{
  vector[N] log_lik;
  {
  vector[N] mu;
  vector[N] gamma;
  for (n in 1:N){
  gamma[n] = bR + bAR*cont_africa[n];
  mu[n] = a + gamma[n] * rugged[n] + bA * cont_africa[n];
  log_lik[n] = normal_lpdf(log_gdp[n] | mu[n], sigma);
  }
  }
}
```

Half cauchy
```{stan output.var=mm2_2}
data {
  int N;
  vector[N] rugged;
  vector[N] cont_africa;
  vector[N] log_gdp;
}
parameters{
  real a;
  real bR;
  real bA;
  real bAR;
  real<lower=0> sigma;
}
model{
  vector[N] gamma;
  vector[N] mu;
  
  for (i in 1:N){
  gamma[i] = bR + bAR * cont_africa[i];
  mu[i] = a + gamma[i] * rugged[i] + bA * cont_africa[i];
  }
  
  target += normal_lpdf(log_gdp | mu, sigma);
  target += normal_lpdf(a | 8, 100);
  target += normal_lpdf(bR | 0, 10);
  target += normal_lpdf(bAR | 0, 10);
  target += normal_lpdf(bA | 0, 10);
  target += cauchy_lpdf(sigma | 0, 0.1);
}
generated quantities{
  vector[N] log_lik;
  {
  vector[N] mu;
  vector[N] gamma;
  for (n in 1:N){
  gamma[n] = bR + bAR*cont_africa[n];
  mu[n] = a + gamma[n] * rugged[n] + bA * cont_africa[n];
  log_lik[n] = normal_lpdf(log_gdp[n] | mu[n], sigma);
  }
  }
}
```

```{r}
fitm_m2_1 <- sampling(mm2_1, data = data_m1, iter = 20000, chains = 4, cores = 4)
fitm_m2_2 <- sampling(mm2_2, data = data_m1, iter = 20000, chains = 4, cores = 4)
```

```{r}
fitm_m2_1
```

```{r}
fitm_m2_2
```

## 8M3
```{r}
fitm_m3_1<- sampling(mm2_2, data = data_m1, iter = 2000, chains = 4, cores = 4, warmup = 500)
```

```{r}
fitm_m3_2<- sampling(mm2_2, data = data_m1, iter = 2000, chains = 4, cores = 4, warmup=100)
```

# Hard

## 8H1

Can get stan code directly from `stancode()` function
```{r}
mp <- map2stan(
  alist(
    a ~ dnorm(0,1),
    b ~ dcauchy(0,1)
  ),
  data=list(y=1),
  start=list(a=0,b=0),
  iter=1e4, warmup=100 , WAIC=FALSE )

stancode(mp)
```

```{stan output.var=mh_1}
data{
    int<lower=1> N;
}
parameters{
    real a;
    real b;
}
model{
    b ~ cauchy( 0 , 1 );
    a ~ normal( 0 , 1 );
}
generated quantities{
    real dev;
    dev = 0;
}
```


## 8H2
Taken from <stan code translation. Adding generated quantities section to do WAIC comparison.
```{stan output.var=mh_2_1}
data {
  int<lower=0> N;
  vector[N] divorce;
  vector[N] median_age_z;
}
parameters {
  real a;
  real bA;
  real<lower=0, upper=10> sigma;
}
model {
  vector[N] mu = a + median_age_z * bA;
  target += normal_lpdf(divorce | mu, sigma);
  target += normal_lpdf(a | 10, 10);
  target += normal_lpdf(bA | 0, 10);
}
generated quantities{
  vector[N] log_lik;
  {
  vector[N] mu;
  for (n in 1:N){
  mu[n] = a + median_age_z[n] * bA;
  log_lik[n] = normal_lpdf(divorce[n] | mu[n], sigma);
  }
  }
}
```

```{stan output.var=mh_2_2}
data {
  int N;
  vector[N] divorce;
  vector[N] marriage_z;
}
parameters {
  real a;
  real bM;
  real<lower=0> sigma;
}
model {
  vector[N] mu = a + marriage_z * bM;
  target += normal_lpdf(divorce | mu, sigma);
  target += normal_lpdf(a | 10, 10);
  target += normal_lpdf(bM | 0, 10);
}
generated quantities{
  vector[N] log_lik;
  {
  vector[N] mu;
  for (n in 1:N){
  mu[n] = a + marriage_z[n] * bM;
  log_lik[n] = normal_lpdf(divorce[n] | mu[n], sigma);
  }
  }
}
```

```{stan output.var=mh_2_3}
data {
  int N;
  vector[N] divorce;
  vector[N] marriage_z;
  vector[N] median_age_z;
}
parameters {
  real a;
  real bA;
  real bM;
  real<lower=0> sigma;
}
model {
  vector[N] mu = a + median_age_z * bA + marriage_z * bM;
  target += normal_lpdf(divorce | mu, sigma);
  target += normal_lpdf(a | 10, 10);
  target += normal_lpdf(bA | 0, 10);
  target += normal_lpdf(bM | 0, 10);
  target += exponential_lpdf(sigma | 1);
}
generated quantities{
  vector[N] log_lik;
  {
  vector[N] mu;
  for (n in 1:N){
  mu[n] = a + median_age_z[n] * bA + marriage_z[n] * bM;;
  log_lik[n] = normal_lpdf(divorce[n] | mu[n], sigma);
  }
  }
}
```

```{r}
data('WaffleDivorce', package = 'rethinking')

WaffleDivorce <- WaffleDivorce %>% 
  mutate(MedianAgeMarriage_z = (MedianAgeMarriage - mean(MedianAgeMarriage)) / sd(MedianAgeMarriage),
         Marriage_z = (Marriage - mean(Marriage)) / sd(Marriage))

data_mh2 <- list(
  N = NROW(WaffleDivorce),
  divorce = WaffleDivorce$Divorce,
  median_age_z = WaffleDivorce$MedianAgeMarriage_z,
  marriage_z = WaffleDivorce$Marriage_z

)

fitm_mh_2_1 <- sampling(mh_2_1, data = data_mh2, iter = 2000, chains = 4, cores = 4)
fitm_mh_2_2 <- sampling(mh_2_2, data = data_mh2, iter = 2000, chains = 4, cores = 4)
fitm_mh_2_3 <- sampling(mh_2_3, data = data_mh2, iter = 2000, chains = 4, cores = 4)

```

```{r warning=FALSE}
# compare models
library(loo)

lok_lik_mh_2_1 <- extract_log_lik(fitm_mh_2_1, merge_chains = FALSE)
waic_mh_2_1 <- waic(lok_lik_mh_2_1)

lok_lik_mh_2_2 <- extract_log_lik(fitm_mh_2_2, merge_chains = FALSE)
waic_mh_2_2<- waic(lok_lik_mh_2_2)

lok_lik_mh_2_3 <- extract_log_lik(fitm_mh_2_3, merge_chains = FALSE)
waic_mh_2_3 <- waic(lok_lik_mh_2_3)

compare_models <- compare(waic_mh_2_1, waic_mh_2_2, waic_mh_2_3)
compare_models
```

## 8H3
```{r}
## R code 8.21
N <- 100                          # number of individuals
height <- rnorm(N,10,2)           # sim total height of each
leg_prop <- runif(N,0.4,0.5)      # leg as proportion of height
leg_left <- leg_prop*height +     # sim left leg as proportion + error
    rnorm( N , 0 , 0.02 )
leg_right <- leg_prop*height +    # sim right leg as proportion + error
    rnorm( N , 0 , 0.02 )
                                  # combine into data frame
d <- data.frame(height,leg_left,leg_right)

## R code 8.22
m5.8s <- map2stan(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + bl*leg_left + br*leg_right ,
        a ~ dnorm( 10 , 100 ) ,
        bl ~ dnorm( 2 , 10 ) ,
        br ~ dnorm( 2 , 10 ) ,
        sigma ~ dcauchy( 0 , 1 )
    ) ,
    data=d, chains=4,
    start=list(a=10,bl=0,br=0,sigma=1))

stancode(m5.8s)
```

```{r}
m5.8s2 <- map2stan(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- a + bl*leg_left + br*leg_right ,
        a ~ dnorm( 10 , 100 ) ,
        bl ~ dnorm( 2 , 10 ) ,
        br ~ dnorm( 2 , 10 ) & T[0,] ,
        sigma ~ dcauchy( 0 , 1 )
    ) ,
    data=d, chains=4,
    start=list(a=10,bl=0,br=0,sigma=1))

stancode(m5.8s2)
```

```{stan output.var=mh_3_1}
data{
    int<lower=1> N;
    real height[N];
    real leg_left[N];
    real leg_right[N];
}
parameters{
    real a;
    real bl;
    real br;
    real<lower=0> sigma;
}
model{
    vector[N] mu;
    sigma ~ cauchy( 0 , 1 );
    br ~ normal( 2 , 10 );
    bl ~ normal( 2 , 10 );
    a ~ normal( 10 , 100 );
    for ( i in 1:N ) {
        mu[i] = a + bl * leg_left[i] + br * leg_right[i];
    }
    height ~ normal( mu , sigma );
}
generated quantities{
 vector[N] log_lik;
  {
    vector[N] mu;
    real dev;
    dev = 0;
    for ( i in 1:N ) {
        mu[i] = a + bl * leg_left[i] + br * leg_right[i];
        log_lik[i] = normal_lpdf(height[i] | mu[i], sigma);
    }
    dev = dev + (-2)*normal_lpdf(height | mu , sigma );
  }
}
```

```{stan output.var=mh_3_2}
data{
    int<lower=1> N;
    real height[N];
    real leg_left[N];
    real leg_right[N];
}
parameters{
    real a;
    real bl;
    real br;
    real<lower=0> sigma;
}
model{
    vector[N] mu;
    sigma ~ cauchy( 0 , 1 );
    br ~ normal( 2 , 10 )T[0, ];
    bl ~ normal( 2 , 10 );
    a ~ normal( 10 , 100 );
    for ( i in 1:N ) {
        mu[i] = a + bl * leg_left[i] + br * leg_right[i];
    }
    height ~ normal( mu , sigma );
}
generated quantities{
 vector[N] log_lik;
  {
    vector[N] mu;
    real dev;
    dev = 0;
    for ( i in 1:N ) {
        mu[i] = a + bl * leg_left[i] + br * leg_right[i];
        log_lik[i] = normal_lpdf(height[i] | mu[i], sigma);
    }
    dev = dev + (-2)*normal_lpdf(height | mu , sigma );
  }
}
```

```{r}
data_mh3<- list(
  N = NROW(d),
  height = d$height,
  leg_left = d$leg_left,
  leg_right = d$leg_right
)

fit_mh_3_1 <- sampling(mh_3_1, data = data_mh3, iter = 2000, chains = 4, cores = 4)
fit_mh_3_2 <- sampling(mh_3_2, data = data_mh3, iter = 2000, chains = 4, cores = 4)

```


```{r}
fit_mh_3_1
```

```{r}
fit_mh_3_2
```

## 8H4
```{r warning=FALSE}
# compare models
library(loo)

lok_lik_mh_3_1 <- extract_log_lik(fit_mh_3_1, merge_chains = FALSE)
waic_mh_3_1 <- waic(lok_lik_mh_3_1)

lok_lik_mh_3_2 <- extract_log_lik(fit_mh_3_2, merge_chains = FALSE)
waic_mh_3_2<- waic(lok_lik_mh_3_2)

compare_models <- compare(waic_mh_3_1, waic_mh_3_2, waic_mh_3_2)
compare_models
```